---
title: "Hands on Exercise 4:  Spatial Weights and Applications"
date: "15-Sep-25"
date-modified: "last-modified"
toc: true
toc-depth: 4
format: html
editor: visual
---

# 1 Overview

In this exercise, I compute and **apply spatial weights** in R. We’ll build contiguity (Queen/Rook) and distance-based neighbours, turn them into weight lists, and use them to create **spatially lagged** indicators for quick diagnostic mapping.

# 2 The Packages

::: panel-tabset
## Package

+----------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| Package                                                                                                                                      | Purpose                                                                                                                 
+==============================================================================================================================================+=============================================================================================================================+
| [**spdep**](https://cran.r-project.org/web/packages/spdep/index.html)                                                                        | To create spatial weights matrix objects and related calculations (e.g. spatially lag attributes).                          |
+----------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| [**sf**](https://cran.r-project.org/web/packages/sf/index.html)                                                                              | For importing, managing, and processing geospatial data                                                                     |
+----------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| [**tidyverse**](https://www.tidyverse.org/)                                                                                                  | A collection of functions for performing data science task such as importing, tidying, wrangling data and visualising data. |
+----------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| [**tmap**](https://cran.r-project.org/web/packages/tmap/)                                                                                    | For plotting thematic maps, such as choropleth and bubble maps                                                              |
+----------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+
| [**DT**](https://rstudio.github.io/DT/), [**knitr**](https://yihui.org/knitr/) and [**kableExtra**](https://haozhu233.github.io/kableExtra/) | For building tables                                                                                                         |
+----------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------+

: {tbl-colwidths="\[15,85\]"}

## Code

```{r}
pacman::p_load(spdep, sf, tmap, tidyverse, knitr, DT, kableExtra)
```
:::

# 3 The Data

We use two datasets:

- Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.
- Hunan_2012.csv: This csv file contains selected Hunan’s local development indicators in 2012.

## 3.1 Loading the data

::: panel-tabset
## Import Geospatial Shapefile

The code chunk below uses [`st_read()`](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

```{r}
#| echo: false

glimpse(hunan)
```

```{r}
#| echo: false
# Setting theme

hunan %>% 
  kable() %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                            fixed_thead = T)
```

## Import Aspatial csv File

Next, we will import *Hunan_2012.csv* into R by using `read_csv()` of **readr** package. The output is R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
#| echo: false
glimpse(hunan2012)
```

```{r}
#| echo: false

hunan2012 %>% 
  kable() %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                            fixed_thead = T)
```

## Perform relational join

The code chunk below will be used to update the attribute table of *hunan*'s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed by using `left_join()` of **dplyr** package.

```{r}
colnames(hunan)
```

```{r}
colnames(hunan2012)
```

```{r}
hunan <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```
:::

## 3.2 Visualising Data

```{r}
#| echo: false
DT::datatable(hunan,
              filter = 'top',
              class = "compact",
              options = list(pageLength = 5, dom = 'tip'))
```

# 4 Visualising Regional Development Indicator

Prepare a basemap and a choropleth map to show the distribution of GDPPC 2012 (GDP per capita) by using `qtm()` of **tmap** package.

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", 
          size=0.3)+
  tm_layout(bg.color = "#fffbe8",
            frame = F)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(bg.color = "#fffbe8",
            frame = F,
            legend.title.size = 0.8,
            legend.text.size = 0.5)

tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

# 5 Computing Contiguity(Adjacency) Spatial Weights

::: {.notebox .note data-latex="note"}
**Contiguity-based**

-   Share common boundary (rooks, bishops, queen's/king's case)
-   Measured using `poly2nb()` of **spdep** package
-   Constructs neighbours list from polygon list
-   This function **builds a neighbours list** based on regions with contiguous boundaries.
-   What this function does is to build a neighbours list based on regions with contiguous boundaries - which means they're in 'contact' and share one or more boundary points. For example, on the Singapore map, the North-East region is contiguous with the North region and the East region.
-   One thing to note is the `queen` argument:
    -   TRUE (default): a single shared boundary point meets the contiguity condition
    -   FALSE: more than one shared point is required
    -   note that more than one shared boundary point does not necessarily mean a shared boundary line

**Distance-based** - Within threshold distance to the nucleus - Measured using `knn2nb()`
:::

## 5.1 Computing contiguity based neighbours

::: panel-tabset \## Queen

The code chunk below is used to compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

::: {.lightbox .light data-latex="light"}
**Interpretation**

-   There are 88 area units in Hunan.
-   Most connected area unit has 11 neighbours.
-   There are 2 area units with only 1 neighbour.
:::

## Rook

The code chunk below is used to compute Rook contiguity weight matrix.

```{r}
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

::: {.lightbox .light data-latex="light"}
**Interpretation**

-   There are 88 area units in Hunan.
-   Most connect area unit has 10 neighbours.
-   There are 2 area units with only 1 neighbour.
:::

### 5.1.1 Display Neighbours for Specific County

For each polygon in our polygon object, `wm_q()` lists all neighboring polygons. For example, to see the neighbors for the first polygon in the object, type:

```{r}
wm_q[[1]]
```

```{r}
wm_r[[1]]
```

Polygon 1 has 5 neighbors. The **numbers** represent the **polygon IDs** as stored in Hunan SpatialPolygonsDataFrame class.

### 5.1.2 Retrieve County Name of Polygon ID=1

```{r}
hunan$County[1]
```

The output reveals that Polygon ID=1 is Anxiang county.

### 5.1.3 Reveal county names of neighboring polygons

```{r}
hunan$NAME_3[c(2,3,4,57,85)]
```

### 5.1.4 Retrieve GDPCC

We can retrieve the GDPPC of these five counties by using the code chunk below.

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

```{r}
nb2 <- wm_r[[1]]
nb2 <- hunan$GDPPC[nb2]
nb2
```

### 5.1.5 Display the complete weight matrix

The printed output above shows that the GDPPC of the five nearest neighbours based on Queen's method are 20981, 34592, 24473, 21311 and 22879 respectively.

You can display the complete weight matrix by using `str()`.

```{r}
str(wm_q)
```

## 5.2 Visualising contiguity weights

A **connectivity graph** draws lines from each unit to its neighbours. Because we have polygons, we’ll use **centroids** as representative points.

### 5.2.1 Get Long & Lat of centroids

::: panel-tabset
## Long

-   To get our longitude values, we map the `st_centroid` function over the geometry column of us.bound and access the longitude value through double bracket notation \[\[\]\] and 1.
-   Allows us to get only the longitude, which is the first value in each centroid.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

## Lat

-   Do the same for latitude with one key difference.
-   We access the second value per each centroid with \[\[2\]\]

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

## Combine Long & Lat

`cbind()` to put longitude and latitude into the same object.

```{r}
coords <- cbind(longitude, latitude)
```

## Check

```{r}
head(coords)
```
:::

### 5.2.2 Plotting contiguity-based neighbours map

```{r fig.width=12, fig.height=8}
#| code-fold: true
#| code-summary: "Show the code"

par(bg = '#fffbe8')

par(mfrow=c(1,2))

plot(hunan$geometry, 
     border="#b89c8c", 
     main="Queen Contiguity")
plot(wm_q, 
     coords, 
     pch = 19, 
     cex = 0.6, 
     add = TRUE, 
     col= "red")

plot(hunan$geometry, 
     border="#b89c8c", 
     main="Rook Contiguity")
plot(wm_r, 
     coords, 
     pch = 19, 
     cex = 0.6, 
     add = TRUE, 
     col = "red")
```

::: {.lightbox .light data-latex="light"}
**Note**

-   There is actually a difference if you look closely at the bottom right of Queen Contiguity plot.
-   There is 2 extra vertices at the bottom right.
:::

# 6 Computing distance-based neighbours

Here we build neighbours by distance bands.

* Use `dnearneigh(coords, d1, d2, longlat = TRUE)` for **banded** distances.
* Use `knearneigh()` + `knn2nb()` for **k-nearest** neighbours.
* `nbdists()` returns edge lengths; with `longlat = TRUE`, distances are in **km** (WGS84).

### 6.2.1 Determine the cut-off distance / upper limit

To determine the upper limit for distance band:

+----------+-----------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| Step     | Task                                                                                                                        | Function                    |
+==========+=============================================================================================================================+=============================+
| 1        | Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other.                  | `knearneigh()` of **spdep** |
+----------+-----------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| 2        | Convert the knn object into a neighbours list of class nb + list of integer vectors containing neighbour region number ids. | `knn2nb()`                  |
+----------+-----------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| 3        | Return the length of neighbour relationship edges.                                                                          | `nbdists()` of **spdep**    |
|          |                                                                                                                             |                             |
|          | The function returns in the units of the coordinates if the coordinates are projected, in km otherwise.                     |                             |
+----------+-----------------------------------------------------------------------------------------------------------------------------+-----------------------------+
| 4        | Remove the list structure of the returned object                                                                            | `unlist()`                  |
+----------+-----------------------------------------------------------------------------------------------------------------------------+-----------------------------+

: {tbl-colwidths="\[5,65,30\]"}

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The summary report shows that the largest first nearest neighbour distance is `61.79` km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

## Computing fixed distance weight matrix

We will compute the distance weight matrix by using `dnearneigh()` function

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

> The "Average number of links: 3.681818" means that, on average, each region is connected to approximately 3.68 neighboring regions based on the specified distance criteria.

We'll display the content of `wm_d62` weight matrix.

::: panel-tabset
### Using `str()`

```{r}
str(wm_d62)
```

### Using `table()` and `card()`

```{r}
table(hunan$County, card(wm_d62))
```
:::

Let us the number of disjoint connected subgraphs in the graph.

```{r}
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

```{r}
table(n_comp$comp.id)
```

### Plotting fixed distance weight matrix

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)
```

The **red lines** show the links of 1st nearest neighbours and the **black lines** show the links of neighbours within the cut-off distance of `62km`.

Alternatively, we can plot both of them next to each other.

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="red", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)
```

## Computing adaptive distance weight matrix

For fixed distance weight matrix, more densely settled areas (usually urban areas) tend to have more neighbours and the less densely settled areas (usually rural counties) tend to have lesser neighbours.

Having many neighbours smoothes the neighbour relationship across more neighbours.

::: panel-tabset
### Control the number of neighbours

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry.

```{r}
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
```

> Notice that each county has at most 6 neighbours.

```{r}
str(knn6)
```

### Plotting distance based neighbours

We can plot the weight matrix using following code:

```{r}
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```
:::

# Weights based on `Inversed Distance Weights` (IDW)

## Step 1: Compute the distance between areas using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

## Step 2: Row-standardised Weights Matrix

Then we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned **equal weight** (`style=“W”`). This is accomplished by assigning the fraction 1/(# of neighbors) to each neighboring county then summing the weighted income values.

While this is the most intuitive way to summaries the neighbors’ values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data.

> For this example, we’ll use `style=“W”` option for simplicity’s sake but more robust options are available, notably `style=“B”`.

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

> `zero.policy = TRUE` option allows for lists of non-neighbors. This should be used with caution since the user may not be aware of missing neighbors in their dataset however, a `zero.policy = FALSE` would return an error.

To see the weight of the first polygon’s eight neighbors:

```{r}
rswm_q$weights[10]
```

Each neighbor is assigned a 0.125 of the total weight. This means that when R computes the average neighboring income values, each neighbor’s income will be multiplied by 0.125 before being tallied.

Using the same method, we can also derive a row standardised distance weight matrix by using following code:

```{r}
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

# Application of Spatial Weight Matrix

In this section, we will learn how to create four different spatial lagged variables, they are:

-   spatial lag with row-standardized weights.
-   spatial lag as a sum of neighbouring values.
-   spatial window average.
-   spatial window sum.

## Spatial lag with row-standardized weights

We will compute the average neighbor GDPPC value for each polygon. These values are often referred to as **spatially lagged values**.

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

Then we can append the spatially lag GDPPC values onto hunan sf data frame.

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

The following table shows the average neighboring income values (stored in the Inc.lag object) for each county.

```{r}
head(hunan)
```

Next, we will plot both the GDPPC and spatial lag GDPPC for comparison.

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

## Spatial lag as a sum of neighboring values

We can calculate spatial lag as a sum of neighboring values by assigning binary weights. This requires us to go back to our neighbors list, then apply a function that will assign binary weights, then we use glist = in the `nb2listw` function to explicitly assign these weights.

We start by applying a function that will assign a value of 1 per each neighbor. This is done with lapply, which we have been using to manipulate the neighbors structure throughout the past notebooks. Basically it applies a function across each value in the neighbors structure.

```{r}
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With the proper weights assigned, we can use `lag.listw` to compute a lag variable from our weight and GDPPC.

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

```{r}
lag_sum
```

Next, we will append the lag_sum GDPPC field into hunan sf data frame.

```{r}
hunan <- left_join(hunan, lag.res)
```

Now, we can plot both the GDPPC and Spatial Lag Sum GDPPC for comparison.

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

## Spatial window average

The spatial window average uses row-standardized weights and includes the diagonal element. To do this in R, we need to go back to the neighbors structure and add the diagonal element before assigning weights.

To add the diagonal element to the neighbour list, we just need to use *include.self()* from **spdep**.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Notice that the Number of nonzero links, Percentage nonzero weights and Average number of links are `536`, `6.921488` and `6.090909` respectively as compared to wm_q of `448`, `5.785124` and `5.090909`.

Let us take a good look at the neighbour list of area \[1\].

```{r}
wm_qs[[1]]
```

Notice that now `[1]` has six neighbours instead of five.

Now we obtain weights with *nb2listw()*

```{r}
wm_qs <- nb2listw(wm_qs)
wm_qs
```

Again, we use *nb2listw()* and *glist()* to explicitly assign weight values.

Lastly, we just need to create the lag variable from our weight structure and GDPPC variable.

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_qs, 
                             hunan$GDPPC)
lag_w_avg_gpdpc
```

Next, we will convert the lag variable listw object into a data.frame by using `as.data.frame()`.

```{r}
lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

Note: The third command line on the code chunk above renames the field names of *lag_wm_q1.res* object into *NAME_3* and *lag_window_avg GDPPC* respectively.

Next, the code chunk below will be used to append *lag_window_avg GDPPC* values onto *hunan* sf data.frame by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan, lag_wm_qs.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of Knitr package is used to prepare a table using the code chunk below.

```{r}
hunan %>%
  select("County", 
         "lag GDPPC", 
         "lag_window_avg GDPPC") %>%
  kable()
```

Lastly, `qtm()` of **tmap** package is used to plot the lag_gdppc and w_ave_gdppc maps next to each other for quick comparison.

```{r}
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)
```

> Note: For more effective comparison, it is advisable to use the core tmap mapping functions.

## Spatial window sum

The spatial window sum is the counter part of the window average, but without using row-standardized weights.

To add the diagonal element to the neighbour list, we just need to use `include.self()` from **spdep**.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Next, we will assign binary weights to the neighbour structure that includes the diagonal element.

```{r}
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]
```

Notice that now `[1]` has six neighbours instead of five.

Again, we use `nb2listw()` and `glist()` to explicitly assign weight values.

```{r}
b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With our new weight structure, we can compute the lag variable with `lag.listw()`.

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

Next, we will convert the lag variable listw object into a data.frame by using `as.data.frame()`.

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

> Note: The second command line on the code chunk above renames the field names of`w_sum_gdppc.res` object into *NAME_3* and `w_sum GDPPC` respectively.

Next, the code chunk below will be used to append *w_sum GDPPC* values onto *hunan* sf data.frame by using `left_join()` of **dplyr** package.

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of **Knitr** package is used to prepare a table using the code chunk below.

```{r}
hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  kable()
```

Lastly, `qtm()` of **tmap** package is used to plot the lag_sum GDPPC and w_sum_gdppc maps next to each other for quick comparison.

```{r}
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp=1, ncol=2)
```
---

### Quick recap (my notes)

* **Queen vs Rook**: Queen is more permissive (corner-touching counts), Rook is stricter (edge-only).
* **Fixed distance vs k-NN**: distance bands can fragment sparse areas; k-NN enforces equal neighbour counts.
* **Window average/sum**: adding the diagonal includes the unit itself; use average (W) vs sum (B) depending on interpretation.
* **Zero policy**: `TRUE` is convenient for isolated units but be explicit about what a 0-lag implies.