---
title: "In class Exercise 8 : Geographically Weighted Predictive Models"
author: "Priya"
date: "October 25, 2025"
date-modified: "last-modified"
execute: 
  eval: false
  echo: true
  message: false
  freeze: true
  warning: false
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    code-overflow: scroll
    number-depth: 4
    code-line-numbers: false
---

## Loading the package

We load the following packages into R environment.

```{r}
pacman::p_load(httr, sf, jsonlite, progress, 
               tmap, rsample, Metrics, tidyverse)
```

## Geocoding HDB resale data 

### Basic version 
```{r}
HDBresale <- read_csv("data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2025-09" & month < "2025-10")
```
### Street name which starts with ST. replace with Saint

```{r}
HDBresale$street_name <- gsub(
  "ST\\.","SAINT",
  HDBresale$street_name)
```


```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block,streetname,sep = " ")
  query <- list("searchVal"= address,
                "returnGeom"= "Y",
                "getAddrDetails"= "N", #make it Y if you want full address
                "pageNum"="1")
  
  res <- GET(base_url, query = query)
  restext<- content(res, as="text")
  
  output<- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)
  
  return(output)
}
```

### Read all the records , return the geocode , insert the value into HDBresale dataframe
```{r}
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0 

for(i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i,4], HDBresale[i,5])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}

```


### Convert HDBresale(13 columns) to spatial data (HDBresale_sf) which is geometry area , make sure it is in SVY 21 projection


```{r}
# Convert to spatial data frame
HDBresale_sf <- HDBresale %>% 
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"),
                         crs = 4326) %>%  # WGS84 for lat/lon coordinates
  st_transform(crs = 3414)  # Transform to SVY21 projection (EPSG:3414)

# Check the result
print(HDBresale_sf)
```


### Preschool data

```{r}
preschool = st_read("data/rawdata/PreSchoolsLocation.kml") %>% 
  st_transform(crs=3414)
```
### All the data are in meters, if we want to search 1km(1000 meters), will return true or false and we calculate the number of T or F. Do the same for childcare,kindergardens,bus stop within 350m

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(prischool_350m = map_int(within_350m,length))
```





