---
title: "Take home Exercise 3 : Predictive Modelling of HDB Resale Prices with Geographically Weighted Methods"
author: "Priya"
date: "November 4, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    code-overflow: scroll
    number-depth: 4
    code-line-numbers: false
---

## Overview

This study develops predictive models for HDB 4-room resale prices in Singapore using geographically weighted methods. We implement and compare multiple modelling approaches including Ordinary Least Squares (OLS), Geographically Weighted Regression (GWR), Random Forest, and Geographically Weighted Random Forest (GRF). The analysis leverages both structural attributes and locational factors to capture the complex spatial patterns in Singapore's public housing market and approaches to identify the most accurate predictive model for housing prices from July to September 2025 based on data from July 2024 to June 2025.

## Background

Singapore's public housing market exhibits unique spatial characteristics due to the island's planned urban development and varying estate maturity. Traditional global regression models often fail to capture local variations in housing price determinants. This study addresses spatial heterogeneity by implementing geographically weighted approaches that allow relationships between price drivers to vary across different neighbourhoods, providing more accurate predictions and deeper insights into local market dynamics.

## Loading Packages

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, tmap, rsample, Metrics, 
               tidyverse, httr, jsonlite, progress, ggplot2, caret, gt,
               corrplot, randomForest, ranger, spatialRF, plotly, ggstatsplot,lubridate,readxl )

```

## Data Preparation

### Load and Explore HDB Resale Data

```{r}
# Load required libraries
library(tidyverse)
library(httr)
library(jsonlite)

#  Load and filter HDB data
cat("=== Loading and Filtering HDB Data ===\n")
HDBresale <- read_csv("data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "5 ROOM") %>%
  filter(month >= "2024-07" & month < "2025-10")

cat("✓ Filtered data:", nrow(HDBresale), "5-room flats from July 2024 to Sept 2025\n")
cat("First 3 records:\n")
print(head(HDBresale, 3))
```
### Geocoding HDB Addresses

```{r}
#  Geocoding HDB Addresses
cat("===  Geocoding HDB Addresses ===\n")

# Create directories if needed
if(!dir.exists("data/rds")) {
  dir.create("data/rds", recursive = TRUE)
  cat("✓ Created data/rds directory\n")
}

# Geocoding function
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext <- content(res, as="text")
  
  output <- fromJSON(restext)  %>% 
    as.data.frame() %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}

# Initialize coordinate columns
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

# Test geocoding with first 5 addresses only
cat("Testing geocoding with first 5 addresses...\n")
for (i in 1:5){
  temp_output <- geocode(HDBresale[i, "block"], HDBresale[i, "street_name"])
  
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
  
  cat("Processed", i, "/ 5 -", 
      HDBresale$block[i], HDBresale$street_name[i], 
      "- Lat:", HDBresale$LATITUDE[i], "Lon:", HDBresale$LONGITUDE[i], "\n")
}

# Display geocoding results
cat("\n=== Geocoding Test Results ===\n")
geocoded_sample <- HDBresale %>% 
  select(block, street_name, LATITUDE, LONGITUDE) %>% 
  head(5)
print(geocoded_sample)

cat("✓ Geocoding test completed with 5 addresses\n")
```

```{r}
#  Full Geocoding with Fixed Data Types
cat("===  Full Geocoding of All Addresses ===\n")

# Improved geocoding function with better error handling
geocode_improved <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  tryCatch({
    res <- GET(base_url, query = query)
    restext <- content(res, as="text", encoding = "UTF-8")
    
    output <- fromJSON(restext)
    
    if (!is.null(output$found) && output$found > 0) {
      # Extract first result coordinates as numeric
      lat <- as.numeric(output$results$LATITUDE[1])
      lon <- as.numeric(output$results$LONGITUDE[1])
      return(list(LATITUDE = lat, LONGITUDE = lon))
    } else {
      return(list(LATITUDE = NA, LONGITUDE = NA))
    }
  }, error = function(e) {
    return(list(LATITUDE = NA, LONGITUDE = NA))
  })
}

# Initialize coordinate columns as numeric
HDBresale$LATITUDE <- NA_real_
HDBresale$LONGITUDE <- NA_real_

# Geocode all addresses (this will take some time)
cat("Starting full geocoding of", nrow(HDBresale), "addresses...\n")
cat("This may take 10-15 minutes. Processing:\n")

for (i in 1:nrow(HDBresale)){
  result <- geocode_improved(HDBresale$block[i], HDBresale$street_name[i])
  
  HDBresale$LATITUDE[i] <- result$LATITUDE
  HDBresale$LONGITUDE[i] <- result$LONGITUDE
  
  # Progress indicator
  if (i %% 100 == 0) {
    cat("Processed", i, "/", nrow(HDBresale), "addresses\n")
  }
}

# Check geocoding results
cat("\n=== Full Geocoding Results ===\n")
successful_geocodes <- sum(!is.na(HDBresale$LATITUDE))
cat("Successful geocodes:", successful_geocodes, "/", nrow(HDBresale), "\n")
cat("Success rate:", round(successful_geocodes / nrow(HDBresale) * 100, 1), "%\n")

# Save the geocoded data
saveRDS(HDBresale, "data/rds/HDBresale_geocoded.rds")
cat("✓ Full geocoded data saved to data/rds/HDBresale_geocoded.rds\n")

# Display final sample
cat("\n=== Final Geocoded Sample ===\n")
final_sample <- HDBresale %>% 
  select(block, street_name, LATITUDE, LONGITUDE) %>% 
  head(8)
print(final_sample)
```



```{r}

#  Data Transformations 
cat("===  Converting Text Fields to Numerical Values ===\n")

# Load the geocoded data
HDBresale <- readRDS("data/rds/HDBresale_geocoded.rds")
cat("Loaded geocoded data:", nrow(HDBresale), "records\n")

# 1. Convert storey_range to storey_mid (EXACT code from example)
cat("\n1. Converting storey_range to storey_mid (exact code)...\n")
storey_mid <- function(x) {
  x %>%
    str_replace_all(" ", "") %>%    # remove spaces
    str_split("TO") %>%             # split into lower and upper floors
    sapply(\(v) mean(as.numeric(v)))  # convert to numeric and take midpoint
}

HDBresale <- HDBresale %>%
  mutate(storey_mid = storey_mid(storey_range)) %>%  # create numeric floor feature
  select(-storey_range)                              # remove the original text column

cat("Storey conversion completed\n")
cat("Sample storey_mid values:\n")
print(head(HDBresale$storey_mid, 10))

# 2. Convert remaining_lease to MONTHS (keeping your requirement)
cat("\n2. Converting remaining_lease to remaining_lease_months...\n")
HDBresale <- HDBresale %>%
  mutate(
    remaining_lease_months = case_when(
      str_detect(remaining_lease, "years") & str_detect(remaining_lease, "months") ~ 
        as.numeric(str_extract(remaining_lease, "\\d+(?= years)")) * 12 +
        as.numeric(str_extract(remaining_lease, "\\d+(?= months)")),
      str_detect(remaining_lease, "years") ~ 
        as.numeric(str_extract(remaining_lease, "\\d+(?= years)")) * 12,
      TRUE ~ NA_real_
    )
  ) %>%
  select(-remaining_lease)  # remove original text column

cat("Remaining lease conversion completed\n")
cat("Sample remaining_lease_months values:\n")
print(head(HDBresale$remaining_lease_months, 10))

# Summary of final dataset
cat("\n=== Final Dataset Summary ===\n")
cat("Total records:", nrow(HDBresale), "\n")
cat("Storey midpoint - Min:", min(HDBresale$storey_mid), "Max:", max(HDBresale$storey_mid), "\n")
cat("Remaining lease months - Min:", min(HDBresale$remaining_lease_months), "Max:", max(HDBresale$remaining_lease_months), "\n")
cat("Coordinates available:", sum(!is.na(HDBresale$LATITUDE)), "/", nrow(HDBresale), "\n")

# Save the fully processed data
saveRDS(HDBresale, "data/rds/HDBresale_processed.rds")
cat("✓ Fully processed data saved to data/rds/HDBresale_processed.rds\n")

# Display final structure
cat("\n=== Final Data Structure ===\n")
print(glimpse(HDBresale))
```
```{r}
#  Convert to Spatial Format
cat("===  Converting to Spatial Format ===\n")

# Load required libraries
library(sf)

# Load the processed data
HDBresale_processed <- readRDS("data/rds/HDBresale_processed.rds")

# Convert to spatial object
HDBresale_sf <- HDBresale_processed %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326  # WGS84
  ) %>%
  st_transform(crs = 3414)  # Convert to SVY21 Singapore projection

cat("✓ Converted to spatial format\n")
cat("Spatial object CRS:", st_crs(HDBresale_sf)$input, "\n")
cat("Number of spatial features:", nrow(HDBresale_sf), "\n")

# Check spatial distribution
cat("\n=== Spatial Distribution ===\n")
bbox <- st_bbox(HDBresale_sf)
cat("Bounding box:\n")
cat("Xmin:", round(bbox$xmin), "Xmax:", round(bbox$xmax), "\n")
cat("Ymin:", round(bbox$ymin), "Ymax:", round(bbox$ymax), "\n")



# Save spatial data
saveRDS(HDBresale_sf, "data/rds/HDBresale_spatial.rds")
cat("✓ Spatial data saved to data/rds/HDBresale_spatial.rds\n")


```

```{r}
#  Train-Test Split and Data Exploration
cat("=== Train-Test Split and Exploratory Analysis ===\n")

# Load the processed data
HDBresale_processed <- readRDS("data/rds/HDBresale_processed.rds")

# Split data into training (July 2024 - June 2025) and test (July 2025 - Sept 2025)
cat("Splitting data into training and test sets...\n")
HDB_train <- HDBresale_processed %>%
  filter(month >= "2024-07" & month <= "2025-06")

HDB_test <- HDBresale_processed %>%
  filter(month >= "2025-07" & month <= "2025-09")

cat("Training set:", nrow(HDB_train), "records (July 2024 - June 2025)\n")
cat("Test set:", nrow(HDB_test), "records (July 2025 - Sept 2025)\n")

# Add dataset identifier for analysis
HDBresale_combined <- bind_rows(
  HDB_train %>% mutate(dataset = "Training"),
  HDB_test %>% mutate(dataset = "Test")
)

# Summary statistics by dataset
cat("\n=== Dataset Summary Statistics ===\n")
summary_stats <- HDBresale_combined %>%
  group_by(dataset) %>%
  summarise(
    n_transactions = n(),
    avg_price = mean(resale_price),
    median_price = median(resale_price),
    min_price = min(resale_price),
    max_price = max(resale_price),
    price_sd = sd(resale_price),
    .groups = 'drop'
  )

print(summary_stats)

# Monthly transaction trends
cat("\n=== Monthly Transaction Trends ===\n")
monthly_summary <- HDBresale_combined %>%
  group_by(month, dataset) %>%
  summarise(
    transaction_count = n(),
    avg_price = mean(resale_price),
    median_price = median(resale_price),
    .groups = 'drop'
  )

print(monthly_summary)

# Save the split datasets
saveRDS(HDB_train, "data/rds/HDB_train.rds")
saveRDS(HDB_test, "data/rds/HDB_test.rds")
saveRDS(HDBresale_combined, "data/rds/HDBresale_combined.rds")

cat("✓ Training and test datasets saved\n")
```

```{r}
#  Data Visualization and Exploration
cat("===  Data Visualization and Exploration ===\n")

# Load required libraries
library(ggplot2)
library(scales)

# Load the combined data
HDBresale_combined <- readRDS("data/rds/HDBresale_combined.rds")

# 1. Monthly transaction trends with price overlay
cat("Creating monthly trends visualization...\n")
monthly_summary <- HDBresale_combined %>%
  group_by(month, dataset) %>%
  summarise(
    transaction_count = n(),
    avg_price = mean(resale_price),
    median_price = median(resale_price),
    .groups = 'drop'
  )

# Plot monthly trends
monthly_trend_plot <- ggplot(monthly_summary, aes(x = month)) +
  geom_bar(aes(y = transaction_count, fill = dataset), 
           stat = "identity", alpha = 0.7, position = "dodge") +
  geom_line(aes(y = avg_price/3000, color = dataset, group = dataset), size = 1) +
  geom_point(aes(y = avg_price/3000, color = dataset), size = 2) +
  scale_y_continuous(
    name = "Number of Transactions",
    sec.axis = sec_axis(~.*3000, name = "Average Price (SGD)", labels = comma)
  ) +
  scale_fill_manual(values = c("Training" = "steelblue", "Test" = "coral")) +
  scale_color_manual(values = c("Training" = "darkblue", "Test" = "darkred")) +
  labs(title = "Monthly Transaction Volume and Price Trends",
       subtitle = "5-Room HDB Resale (July 2024 - Sept 2025)",
       x = "Month",
       fill = "Dataset",
       color = "Dataset") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")

print(monthly_trend_plot)

# 2. Price distribution by dataset
cat("Creating price distribution visualization...\n")
price_hist <- ggplot(HDBresale_combined, aes(x = resale_price, fill = dataset)) +
  geom_histogram(alpha = 0.7, bins = 50, position = "identity") +
  scale_x_continuous(labels = dollar, breaks = seq(0, 2000000, 200000)) +
  scale_fill_manual(values = c("Training" = "steelblue", "Test" = "coral")) +
  labs(title = "Distribution of HDB 5-Room Resale Prices",
       subtitle = "Training vs Test Periods",
       x = "Resale Price (SGD)",
       y = "Count",
       fill = "Dataset") +
  theme_minimal() +
  facet_wrap(~dataset, ncol = 1, scales = "free_y")

print(price_hist)

# 3. Structural factors analysis
cat("Analyzing structural factors...\n")
structural_plot <- ggplot(HDBresale_combined, aes(x = floor_area_sqm, y = resale_price, color = dataset)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = dollar) +
  scale_color_manual(values = c("Training" = "steelblue", "Test" = "coral")) +
  labs(title = "Floor Area vs Resale Price",
       x = "Floor Area (sqm)",
       y = "Resale Price (SGD)",
       color = "Dataset") +
  theme_minimal()

print(structural_plot)

# Display key insights
cat("\n=== Key Insights ===\n")
cat("• Training set:", nrow(filter(HDBresale_combined, dataset == "Training")), "transactions\n")
cat("• Test set:", nrow(filter(HDBresale_combined, dataset == "Test")), "transactions\n")
cat("• Price increase from training to test: +SGD", 
    round((784807.5 - 758386.0), 0), "(+3.5%)\n")
cat("• Average monthly transactions in training:", round(6264/12, 0), "\n")
cat("• Average monthly transactions in test:", round(1607/3, 0), "\n")

# Save visualizations
ggsave("data/rds/monthly_trends_plot.png", monthly_trend_plot, width = 12, height = 6, dpi = 300)
ggsave("data/rds/price_distribution_plot.png", price_hist, width = 10, height = 8, dpi = 300)
ggsave("data/rds/structural_factors_plot.png", structural_plot, width = 10, height = 6, dpi = 300)

cat("✓ Visualizations saved\n")
```


## Location factors

### Proxomity to CBD

```{r}
# Proximity to CBD (Using Exact Coordinates)
cat("===  Proximity to CBD ===\n")

# Load required libraries
library(sf)

# Load the training data and convert to spatial
HDB_train_sf <- readRDS("data/rds/HDB_train.rds") %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326
  ) %>%
  st_transform(crs = 3414)  # SVY21 Singapore projection

cat("Training data converted to spatial format\n")
cat("Number of training records:", nrow(HDB_train_sf), "\n")

# Define CBD coordinates (EXACT from the code)
cat("Calculating distance to CBD...\n")
cbd_lon <- 103.85    # Exact from code
cbd_lat <- 1.28967   # Exact from code

# Create CBD point and transform to SVY21 (EXACT from code)
cbd_wgs84 <- st_sfc(st_point(c(cbd_lon, cbd_lat)), crs = 4326)
cbd_3414 <- st_transform(cbd_wgs84, 3414)

# Compute straight-line distance to CBD (meters and km - EXACT from code)
HDB_train_sf <- HDB_train_sf %>%
  mutate(
    proximity_to_cbd_m = as.numeric(st_distance(geometry, cbd_3414)[, 1]),
    proximity_to_cbd_km = proximity_to_cbd_m / 1000
  )

# Remove meters column (as in the example code)
HDB_train_sf <- HDB_train_sf %>%
  select(-proximity_to_cbd_m)

# Display results
cat("✓ CBD proximity calculated\n")
cat("\n=== CBD Proximity Summary ===\n")
cat("Minimum distance to CBD:", round(min(HDB_train_sf$proximity_to_cbd_km), 2), "km\n")
cat("Maximum distance to CBD:", round(max(HDB_train_sf$proximity_to_cbd_km), 2), "km\n")
cat("Average distance to CBD:", round(mean(HDB_train_sf$proximity_to_cbd_km), 2), "km\n")
cat("Median distance to CBD:", round(median(HDB_train_sf$proximity_to_cbd_km), 2), "km\n")

# Show sample of results (closest to CBD)
cat("\n=== Closest Properties to CBD ===\n")
closest_cbd <- HDB_train_sf %>%
  select(block, street_name, town, proximity_to_cbd_km) %>%
  st_drop_geometry() %>%
  arrange(proximity_to_cbd_km) %>%
  head(10)
print(closest_cbd)

# Show sample of results (furthest from CBD)
cat("\n=== Furthest Properties from CBD ===\n")
furthest_cbd <- HDB_train_sf %>%
  select(block, street_name, town, proximity_to_cbd_km) %>%
  st_drop_geometry() %>%
  arrange(desc(proximity_to_cbd_km)) %>%
  head(10)
print(furthest_cbd)

# Save with CBD proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_cbd.rds")
cat("✓ Training data with CBD proximity saved\n")

# Quick visualization
cat("\nCreating CBD distance distribution plot...\n")
library(ggplot2)

cbd_plot <- ggplot(HDB_train_sf %>% st_drop_geometry(), aes(x = proximity_to_cbd_km)) +
  geom_histogram(fill = "steelblue", alpha = 0.7, bins = 30) +
  geom_vline(aes(xintercept = mean(proximity_to_cbd_km)), 
             color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = mean(HDB_train_sf$proximity_to_cbd_km) + 2, 
           y = 300, label = paste("Mean:", round(mean(HDB_train_sf$proximity_to_cbd_km), 1), "km"), 
           color = "red") +
  labs(title = "Distribution of Distance to CBD",
       subtitle = "HDB 5-Room Flats (Training Set) - Using Exact CBD Coordinates",
       x = "Distance to CBD (km)",
       y = "Number of Properties") +
  theme_minimal()

print(cbd_plot)
```
### Proximity to eldercare

```{r}
#Proximity to Eldercare
cat("===  Proximity to Eldercare ===\n")

# Load the training data with CBD proximity
HDB_train_sf <- readRDS("data/rds/HDB_train_with_cbd.rds")

# Load eldercare data
cat("Loading eldercare data...\n")
if(file.exists("data/rawdata/EldercareServicesKML.kml")) {
  
  eldercare <- st_read("data/rawdata/EldercareServicesKML.kml", quiet = TRUE) %>%
    st_transform(crs = 3414)
  st_crs(eldercare)  # check the CRS
  
  cat("✓ Eldercare data loaded\n")
} else {
  stop("EldercareServicesKML.kml not found!")
}

# Check eldercare data
cat("Eldercare facilities found:", nrow(eldercare), "\n")
cat("Eldercare CRS:", st_crs(eldercare)$input, "\n")

# Calculate distance to nearest eldercare facility (in kilometers)
cat("Calculating distance to nearest eldercare...\n")
nearest_eldercare_idx <- st_nearest_feature(HDB_train_sf, eldercare)
dist_eldercare_km <- as.numeric(st_distance(HDB_train_sf, eldercare[nearest_eldercare_idx, ], by_element = TRUE)) / 1000

# Count eldercare facilities within 350m
cat("Counting eldercare facilities within 350m...\n")
within_350m_eldercare <- st_is_within_distance(HDB_train_sf, eldercare, dist = 350)
count_eldercare_350m <- lengths(within_350m_eldercare)

# Add both proximity measures to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(
    dist_eldercare_km = dist_eldercare_km,
    eldercare_350m = count_eldercare_350m
  )

# Display results
cat("✓ Eldercare proximity analysis completed\n")
cat("\n=== Eldercare Proximity Summary ===\n")
cat("Distance to nearest eldercare:\n")
cat("  Min:", round(min(HDB_train_sf$dist_eldercare_km), 2), "km\n")
cat("  Max:", round(max(HDB_train_sf$dist_eldercare_km), 2), "km\n")
cat("  Mean:", round(mean(HDB_train_sf$dist_eldercare_km), 2), "km\n")
cat("  Median:", round(median(HDB_train_sf$dist_eldercare_km), 2), "km\n")

cat("\nEldercare facilities within 350m:\n")
cat("  Properties with 0 facilities:", sum(HDB_train_sf$eldercare_350m == 0), "\n")
cat("  Properties with 1+ facilities:", sum(HDB_train_sf$eldercare_350m >= 1), "\n")
cat("  Properties with 2+ facilities:", sum(HDB_train_sf$eldercare_350m >= 2), "\n")
cat("  Maximum facilities within 350m:", max(HDB_train_sf$eldercare_350m), "\n")

# Show sample of results
cat("\n=== Sample Eldercare Proximity ===\n")
eldercare_sample <- HDB_train_sf %>%
  select(block, street_name, town, dist_eldercare_km, eldercare_350m) %>%
  st_drop_geometry() %>%
  arrange(dist_eldercare_km) %>%
  head(10)
print(eldercare_sample)

# Save with eldercare proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_eldercare.rds")
cat("✓ Training data with eldercare proximity saved\n")


```
### Proximity to Hawker Centres

```{r}
#  Proximity to Hawker Centres
cat("===  Proximity to Hawker Centres ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_eldercare.rds")

# Load hawker centres data
cat("Loading hawker centres data...\n")
if(file.exists("data/rawdata/HawkerCentresKML.kml")) {
  hawker <- st_read("data/rawdata/HawkerCentresKML.kml", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ Hawker centres data loaded from KML\n")
} else if(file.exists("data/rawdata/HawkerCentresGEOJSON.geojson")) {
  hawker <- st_read("data/rawdata/HawkerCentresGEOJSON.geojson", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ Hawker centres data loaded from GeoJSON\n")
} else {
  stop("Hawker centres data not found!")
}

# Check hawker data
cat("Hawker centres found:", nrow(hawker), "\n")
cat("Hawker centres CRS:", st_crs(hawker)$input, "\n")

# Calculate distance to nearest hawker centre (in kilometers)
cat("Calculating distance to nearest hawker centre...\n")
nearest_hawker_idx <- st_nearest_feature(HDB_train_sf, hawker)
dist_hawker_km <- as.numeric(st_distance(HDB_train_sf, hawker[nearest_hawker_idx, ], by_element = TRUE)) / 1000

# Add distance to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(dist_hawker_km = dist_hawker_km)

# Display results
cat("✓ Hawker centres proximity analysis completed\n")
cat("\n=== Hawker Centres Proximity Summary ===\n")
cat("Distance to nearest hawker centre:\n")
cat("  Min:", round(min(HDB_train_sf$dist_hawker_km), 2), "km\n")
cat("  Max:", round(max(HDB_train_sf$dist_hawker_km), 2), "km\n")
cat("  Mean:", round(mean(HDB_train_sf$dist_hawker_km), 2), "km\n")
cat("  Median:", round(median(HDB_train_sf$dist_hawker_km), 2), "km\n")

# Show distribution statistics
cat("\nDistance distribution:\n")
quantiles <- quantile(HDB_train_sf$dist_hawker_km, probs = c(0.1, 0.25, 0.5, 0.75, 0.9))
cat("  10th percentile:", round(quantiles[1], 2), "km\n")
cat("  25th percentile:", round(quantiles[2], 2), "km\n")
cat("  50th percentile:", round(quantiles[3], 2), "km\n")
cat("  75th percentile:", round(quantiles[4], 2), "km\n")
cat("  90th percentile:", round(quantiles[5], 2), "km\n")

# Show sample of results
cat("\n=== Sample Hawker Centres Proximity ===\n")
hawker_sample <- HDB_train_sf %>%
  select(block, street_name, town, dist_hawker_km) %>%
  st_drop_geometry() %>%
  arrange(dist_hawker_km) %>%
  head(10)
print(hawker_sample)

# Save with hawker centres proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_hawker.rds")
cat("✓ Training data with hawker centres proximity saved\n")

```
### Proximity to MRT Stations

```{r}
#  Proximity to MRT Stations
cat("=== Proximity to MRT Stations ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_hawker.rds")

# Load MRT stations data
cat("Loading MRT stations data...\n")
if(file.exists("data/rawdata/LTAMRTStationExitKML.kml")) {
  mrt <- st_read("data/rawdata/LTAMRTStationExitKML.kml", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ MRT stations data loaded from KML\n")
} else {
  stop("MRT stations data not found!")
}

# Check MRT data
cat("MRT stations found:", nrow(mrt), "\n")
cat("MRT stations CRS:", st_crs(mrt)$input, "\n")

# Calculate distance to nearest MRT station (in kilometers)
cat("Calculating distance to nearest MRT station...\n")
nearest_mrt_idx <- st_nearest_feature(HDB_train_sf, mrt)
dist_mrt_km <- as.numeric(st_distance(HDB_train_sf, mrt[nearest_mrt_idx, ], by_element = TRUE)) / 1000

# Add distance to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(dist_mrt_km = dist_mrt_km)

# Display results
cat("✓ MRT stations proximity analysis completed\n")
cat("\n=== MRT Stations Proximity Summary ===\n")
cat("Distance to nearest MRT station:\n")
cat("  Min:", round(min(HDB_train_sf$dist_mrt_km), 2), "km\n")
cat("  Max:", round(max(HDB_train_sf$dist_mrt_km), 2), "km\n")
cat("  Mean:", round(mean(HDB_train_sf$dist_mrt_km), 2), "km\n")
cat("  Median:", round(median(HDB_train_sf$dist_mrt_km), 2), "km\n")

# Show distribution statistics
cat("\nDistance distribution:\n")
quantiles <- quantile(HDB_train_sf$dist_mrt_km, probs = c(0.1, 0.25, 0.5, 0.75, 0.9))
cat("  10th percentile:", round(quantiles[1], 2), "km\n")
cat("  25th percentile:", round(quantiles[2], 2), "km\n")
cat("  50th percentile:", round(quantiles[3], 2), "km\n")
cat("  75th percentile:", round(quantiles[4], 2), "km\n")
cat("  90th percentile:", round(quantiles[5], 2), "km\n")

# Categorize MRT proximity
HDB_train_sf <- HDB_train_sf %>%
  mutate(mrt_proximity = case_when(
    dist_mrt_km <= 0.5 ~ "Very Close (≤0.5km)",
    dist_mrt_km <= 1.0 ~ "Close (0.5-1km)",
    dist_mrt_km <= 2.0 ~ "Moderate (1-2km)",
    TRUE ~ "Far (>2km)"
  ))

cat("\nMRT Proximity Categories:\n")
mrt_categories <- HDB_train_sf %>% st_drop_geometry() %>% count(mrt_proximity)
print(mrt_categories)

# Show sample of results
cat("\n=== Sample MRT Stations Proximity ===\n")
mrt_sample <- HDB_train_sf %>%
  select(block, street_name, town, dist_mrt_km, mrt_proximity) %>%
  st_drop_geometry() %>%
  arrange(dist_mrt_km) %>%
  head(10)
print(mrt_sample)

# Save with MRT proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_mrt.rds")
cat("✓ Training data with MRT proximity saved\n")

```
### Proximity to Parks

```{r}
#  Proximity to Parks
cat("=== Proximity to Parks ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_mrt.rds")

# Load parks data
cat("Loading parks data...\n")
if(file.exists("data/rawdata/ParkFacilities.geojson")) {
  parks <- st_read("data/rawdata/ParkFacilities.geojson", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ Parks data loaded from GeoJSON\n")
} else {
  stop("ParkFacilities.geojson not found!")
}

# Check parks data
cat("Park facilities found:", nrow(parks), "\n")
cat("Parks CRS:", st_crs(parks)$input, "\n")

# Calculate distance to nearest park (in kilometers)
cat("Calculating distance to nearest park...\n")
nearest_park_idx <- st_nearest_feature(HDB_train_sf, parks)
dist_park_km <- as.numeric(st_distance(HDB_train_sf, parks[nearest_park_idx, ], by_element = TRUE)) / 1000

# Add distance to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(dist_park_km = dist_park_km)

# Display results
cat("✓ Parks proximity analysis completed\n")
cat("\n=== Parks Proximity Summary ===\n")
cat("Distance to nearest park:\n")
cat("  Min:", round(min(HDB_train_sf$dist_park_km), 2), "km\n")
cat("  Max:", round(max(HDB_train_sf$dist_park_km), 2), "km\n")
cat("  Mean:", round(mean(HDB_train_sf$dist_park_km), 2), "km\n")
cat("  Median:", round(median(HDB_train_sf$dist_park_km), 2), "km\n")

# Show distribution statistics
cat("\nDistance distribution:\n")
quantiles <- quantile(HDB_train_sf$dist_park_km, probs = c(0.1, 0.25, 0.5, 0.75, 0.9))
cat("  10th percentile:", round(quantiles[1], 2), "km\n")
cat("  25th percentile:", round(quantiles[2], 2), "km\n")
cat("  50th percentile:", round(quantiles[3], 2), "km\n")
cat("  75th percentile:", round(quantiles[4], 2), "km\n")
cat("  90th percentile:", round(quantiles[5], 2), "km\n")

# Categorize park proximity
HDB_train_sf <- HDB_train_sf %>%
  mutate(park_proximity = case_when(
    dist_park_km <= 0.3 ~ "Very Close (≤0.3km)",
    dist_park_km <= 0.6 ~ "Close (0.3-0.6km)",
    dist_park_km <= 1.0 ~ "Moderate (0.6-1km)",
    TRUE ~ "Far (>1km)"
  ))

cat("\nPark Proximity Categories:\n")
park_categories <- HDB_train_sf %>% st_drop_geometry() %>% count(park_proximity)
print(park_categories)

# Show sample of results
cat("\n=== Sample Parks Proximity ===\n")
park_sample <- HDB_train_sf %>%
  select(block, street_name, town, dist_park_km, park_proximity) %>%
  st_drop_geometry() %>%
  arrange(dist_park_km) %>%
  head(10)
print(park_sample)

# Save with parks proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_parks.rds")
cat("✓ Training data with parks proximity saved\n")

# Quick visualization
cat("\nCreating parks proximity plots...\n")
library(patchwork)

# Plot 1: Distance distribution
plot1 <- ggplot(HDB_train_sf %>% st_drop_geometry(), aes(x = dist_park_km)) +
  geom_histogram(fill = "darkgreen", alpha = 0.7, bins = 30) +
  geom_vline(aes(xintercept = median(dist_park_km)), 
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distance to Nearest Park",
       x = "Distance (km)", y = "Count") +
  theme_minimal()

# Plot 2: Proximity categories
plot2 <- ggplot(HDB_train_sf %>% st_drop_geometry(), aes(x = park_proximity)) +
  geom_bar(fill = "lightgreen", alpha = 0.7) +
  labs(title = "Park Proximity Categories",
       x = "Proximity Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(plot1 + plot2)


```
### Proximity to Good Primary Schools

```{r}
#  Proximity to Good Primary Schools
cat("===  Proximity to Good Primary Schools ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_parks.rds")

# Check if we have primary schools data
cat("Looking for primary schools data...\n")
if(file.exists("data/rawdata/Generalinformationofschools.xlsx")) {
  # Read Excel file with primary schools
  library(readxl)
  primary_schools <- read_excel("data/rawdata/Generalinformationofschools.xlsx") %>%
    filter(!is.na(`Postal Code`)) %>%
    mutate(postal_code = str_pad(as.character(`Postal Code`), 6, pad = "0"))
  cat("✓ Primary schools data loaded from Excel\n")
} else {
  # Create a sample of top primary schools if no data exists
  cat("No primary schools data found. Creating sample data...\n")
  primary_schools <- tibble(
    school_name = c("Raffles Girls' Primary School", "Nanyang Primary School", "Tao Nan School", 
                   "Rosyth School", "Henry Park Primary School", "Nan Hua Primary School",
                   "Catholic High School (Primary)", "St. Hilda's Primary School",
                   "Ai Tong School", "CHIJ St. Nicholas Girls' School (Primary)"),
    postal_code = c("258499", "678633", "424242", "538527", "279411", "599634", 
                   "569761", "489338", "569930", "569961")
  )
}

# Geocode primary schools using OneMap API
cat("Geocoding primary schools...\n")
geocode_school <- function(postal_code) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  query <- list("searchVal" = postal_code,
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  tryCatch({
    res <- GET(base_url, query = query, timeout(10))
    if (status_code(res) == 200) {
      content_data <- content(res, as = "parsed")
      if (!is.null(content_data$found) && content_data$found > 0) {
        first_result <- content_data$results[[1]]
        return(list(
          LATITUDE = as.numeric(first_result$LATITUDE),
          LONGITUDE = as.numeric(first_result$LONGITUDE),
          status = "success"
        ))
      }
    }
    return(list(LATITUDE = NA, LONGITUDE = NA, status = "no_results"))
  }, error = function(e) {
    return(list(LATITUDE = NA, LONGITUDE = NA, status = "error"))
  })
}

# Geocode schools (limited to first 20 for speed)
primary_schools$LATITUDE <- NA
primary_schools$LONGITUDE <- NA

cat("Geocoding", nrow(primary_schools), "primary schools...\n")
for(i in 1:min(20, nrow(primary_schools))) {
  result <- geocode_school(primary_schools$postal_code[i])
  primary_schools$LATITUDE[i] <- result$LATITUDE
  primary_schools$LONGITUDE[i] <- result$LONGITUDE
}

# Convert to spatial object
primary_schools_sf <- primary_schools %>%
  filter(!is.na(LATITUDE)) %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326
  ) %>%
  st_transform(crs = 3414)

cat("Successfully geocoded", nrow(primary_schools_sf), "primary schools\n")

# Calculate distance to nearest good primary school (in kilometers)
cat("Calculating distance to nearest good primary school...\n")
nearest_school_idx <- st_nearest_feature(HDB_train_sf, primary_schools_sf)
dist_school_km <- as.numeric(st_distance(HDB_train_sf, primary_schools_sf[nearest_school_idx, ], by_element = TRUE)) / 1000

# Count primary schools within 1km
cat("Counting primary schools within 1km...\n")
within_1km_schools <- st_is_within_distance(HDB_train_sf, primary_schools_sf, dist = 1000)
count_schools_1km <- lengths(within_1km_schools)

# Add both proximity measures to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(
    dist_good_ps_km = dist_school_km,
    primary_schools_1km = count_schools_1km
  )

# Display results
cat("✓ Good primary schools proximity analysis completed\n")
cat("\n=== Good Primary Schools Proximity Summary ===\n")
cat("Distance to nearest good primary school:\n")
cat("  Min:", round(min(HDB_train_sf$dist_good_ps_km), 2), "km\n")
cat("  Max:", round(max(HDB_train_sf$dist_good_ps_km), 2), "km\n")
cat("  Mean:", round(mean(HDB_train_sf$dist_good_ps_km), 2), "km\n")
cat("  Median:", round(median(HDB_train_sf$dist_good_ps_km), 2), "km\n")

cat("\nPrimary schools within 1km:\n")
cat("  Properties with 0 schools:", sum(HDB_train_sf$primary_schools_1km == 0), "\n")
cat("  Properties with 1+ schools:", sum(HDB_train_sf$primary_schools_1km >= 1), "\n")
cat("  Properties with 2+ schools:", sum(HDB_train_sf$primary_schools_1km >= 2), "\n")
cat("  Maximum schools within 1km:", max(HDB_train_sf$primary_schools_1km), "\n")

# Show sample of results
cat("\n=== Sample Good Primary Schools Proximity ===\n")
school_sample <- HDB_train_sf %>%
  select(block, street_name, town, dist_good_ps_km, primary_schools_1km) %>%
  st_drop_geometry() %>%
  arrange(dist_good_ps_km) %>%
  head(10)
print(school_sample)

# Save with primary schools proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_schools.rds")
cat("✓ Training data with primary schools proximity saved\n")
```

### Proximity to Shopping Malls

```{r}
#  Proximity to Shopping Malls
cat("===  Proximity to Shopping Malls ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_schools.rds")

# Load shopping malls data
cat("Loading shopping malls data...\n")
if(file.exists("data/rawdata/shopping_mall_coordinates.xlsx")) {
  library(readxl)
  malls <- read_excel("data/rawdata/shopping_mall_coordinates.xlsx") %>%
    filter(!is.na(postal_code)) %>%
    mutate(postal_code = str_pad(as.character(postal_code), 6, pad = "0"))
  cat("✓ Shopping malls data loaded from Excel\n")
} else {
  # Create sample shopping malls data
  cat("No shopping malls data found. Creating sample data...\n")
  malls <- tibble(
    mall_name = c("VivoCity", "Junction 8", "NEX", "Tampines Mall", "Jurong Point", 
                 "Northpoint City", "Bedok Mall", "Causeway Point", "Lot One", "Westgate"),
    postal_code = c("098585", "579837", "556083", "529457", "648886", 
                   "768098", "467360", "738099", "677742", "608532")
  )
}

# Geocode shopping malls using OneMap API
cat("Geocoding shopping malls...\n")
geocode_mall <- function(postal_code) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  query <- list("searchVal" = postal_code,
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  tryCatch({
    res <- GET(base_url, query = query, timeout(10))
    if (status_code(res) == 200) {
      content_data <- content(res, as = "parsed")
      if (!is.null(content_data$found) && content_data$found > 0) {
        first_result <- content_data$results[[1]]
        return(list(
          LATITUDE = as.numeric(first_result$LATITUDE),
          LONGITUDE = as.numeric(first_result$LONGITUDE),
          status = "success"
        ))
      }
    }
    return(list(LATITUDE = NA, LONGITUDE = NA, status = "no_results"))
  }, error = function(e) {
    return(list(LATITUDE = NA, LONGITUDE = NA, status = "error"))
  })
}

# Geocode malls
malls$LATITUDE <- NA
malls$LONGITUDE <- NA

cat("Geocoding", nrow(malls), "shopping malls...\n")
for(i in 1:nrow(malls)) {
  result <- geocode_mall(malls$postal_code[i])
  malls$LATITUDE[i] <- result$LATITUDE
  malls$LONGITUDE[i] <- result$LONGITUDE
}

# Convert to spatial object
malls_sf <- malls %>%
  filter(!is.na(LATITUDE)) %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326
  ) %>%
  st_transform(crs = 3414)

cat("Successfully geocoded", nrow(malls_sf), "shopping malls\n")

# Calculate distance to nearest shopping mall (in kilometers)
cat("Calculating distance to nearest shopping mall...\n")
nearest_mall_idx <- st_nearest_feature(HDB_train_sf, malls_sf)
dist_mall_km <- as.numeric(st_distance(HDB_train_sf, malls_sf[nearest_mall_idx, ], by_element = TRUE)) / 1000

# Count shopping malls within 1km
cat("Counting shopping malls within 1km...\n")
within_1km_malls <- st_is_within_distance(HDB_train_sf, malls_sf, dist = 1000)
count_malls_1km <- lengths(within_1km_malls)

# Add both proximity measures to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(
    dist_mall_km = dist_mall_km,
    malls_1km = count_malls_1km
  )

# Display results
cat("✓ Shopping malls proximity analysis completed\n")
cat("\n=== Shopping Malls Proximity Summary ===\n")
cat("Distance to nearest shopping mall:\n")
cat("  Min:", round(min(HDB_train_sf$dist_mall_km), 2), "km\n")
cat("  Max:", round(max(HDB_train_sf$dist_mall_km), 2), "km\n")
cat("  Mean:", round(mean(HDB_train_sf$dist_mall_km), 2), "km\n")
cat("  Median:", round(median(HDB_train_sf$dist_mall_km), 2), "km\n")

cat("\nShopping malls within 1km:\n")
cat("  Properties with 0 malls:", sum(HDB_train_sf$malls_1km == 0), "\n")
cat("  Properties with 1+ malls:", sum(HDB_train_sf$malls_1km >= 1), "\n")
cat("  Properties with 2+ malls:", sum(HDB_train_sf$malls_1km >= 2), "\n")
cat("  Maximum malls within 1km:", max(HDB_train_sf$malls_1km), "\n")

# Show sample of results
cat("\n=== Sample Shopping Malls Proximity ===\n")
mall_sample <- HDB_train_sf %>%
  select(block, street_name, town, dist_mall_km, malls_1km) %>%
  st_drop_geometry() %>%
  arrange(dist_mall_km) %>%
  head(10)
print(mall_sample)

# Save with shopping malls proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_malls.rds")
cat("✓ Training data with shopping malls proximity saved\n")


```
### Proximity to Supermarkets

```{r}
#  Proximity to Supermarkets
cat("===  Proximity to Supermarkets ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_malls.rds")

# Load supermarkets data
cat("Loading supermarkets data...\n")
if(file.exists("data/rawdata/SupermarketsGEOJSON.geojson")) {
  supermarkets <- st_read("data/rawdata/SupermarketsGEOJSON.geojson", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ Supermarkets data loaded from GeoJSON\n")
} else {
  stop("SupermarketsGEOJSON.geojson not found!")
}

# Check supermarkets data
cat("Supermarkets found:", nrow(supermarkets), "\n")
cat("Supermarkets CRS:", st_crs(supermarkets)$input, "\n")

# Calculate distance to nearest supermarket (in kilometers)
cat("Calculating distance to nearest supermarket...\n")
nearest_supermarket_idx <- st_nearest_feature(HDB_train_sf, supermarkets)
dist_supermarket_km <- as.numeric(st_distance(HDB_train_sf, supermarkets[nearest_supermarket_idx, ], by_element = TRUE)) / 1000

# Add distance to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(dist_supermarket_km = dist_supermarket_km)

# Display results
cat("✓ Supermarkets proximity analysis completed\n")
cat("\n=== Supermarkets Proximity Summary ===\n")
cat("Distance to nearest supermarket:\n")
cat("  Min:", round(min(HDB_train_sf$dist_supermarket_km), 2), "km\n")
cat("  Max:", round(max(HDB_train_sf$dist_supermarket_km), 2), "km\n")
cat("  Mean:", round(mean(HDB_train_sf$dist_supermarket_km), 2), "km\n")
cat("  Median:", round(median(HDB_train_sf$dist_supermarket_km), 2), "km\n")

# Show distribution statistics
cat("\nDistance distribution:\n")
quantiles <- quantile(HDB_train_sf$dist_supermarket_km, probs = c(0.1, 0.25, 0.5, 0.75, 0.9))
cat("  10th percentile:", round(quantiles[1], 2), "km\n")
cat("  25th percentile:", round(quantiles[2], 2), "km\n")
cat("  50th percentile:", round(quantiles[3], 2), "km\n")
cat("  75th percentile:", round(quantiles[4], 2), "km\n")
cat("  90th percentile:", round(quantiles[5], 2), "km\n")

# Show sample of results
cat("\n=== Sample Supermarkets Proximity ===\n")
supermarket_sample <- HDB_train_sf %>%
  select(block, street_name, town, dist_supermarket_km) %>%
  st_drop_geometry() %>%
  arrange(dist_supermarket_km) %>%
  head(10)
print(supermarket_sample)

# Save with supermarkets proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_supermarkets.rds")
cat("✓ Training data with supermarkets proximity saved\n")


```
### Number of Kindergartens within 350m

```{r}
#  Kindergartens within 350m
cat("=== Kindergartens within 350m ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_supermarkets.rds")

# Load kindergartens data
cat("Loading kindergartens data...\n")
if(file.exists("data/rawdata/Kindergartens.geojson")) {
  kindergartens <- st_read("data/rawdata/Kindergartens.geojson", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ Kindergartens data loaded from GeoJSON\n")
} else if(file.exists("data/rawdata/PreSchoolsLocation.kml")) {
  kindergartens <- st_read("data/rawdata/PreSchoolsLocation.kml", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ Kindergartens data loaded from KML (using preschools)\n")
} else {
  stop("Kindergartens data not found!")
}

# Check kindergartens data
cat("Kindergartens found:", nrow(kindergartens), "\n")
cat("Kindergartens CRS:", st_crs(kindergartens)$input, "\n")

# Count kindergartens within 350m of each HDB
cat("Counting kindergartens within 350m...\n")
within_350m_kg <- st_is_within_distance(HDB_train_sf, kindergartens, dist = 350)
count_kg_350m <- lengths(within_350m_kg)

# Add count to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(kindergartens_350m = count_kg_350m)

# Display results
cat("✓ Kindergartens proximity analysis completed\n")
cat("\n=== Kindergartens within 350m Summary ===\n")
cat("Properties with 0 kindergartens:", sum(HDB_train_sf$kindergartens_350m == 0), "\n")
cat("Properties with 1+ kindergartens:", sum(HDB_train_sf$kindergartens_350m >= 1), "\n")
cat("Properties with 2+ kindergartens:", sum(HDB_train_sf$kindergartens_350m >= 2), "\n")
cat("Properties with 5+ kindergartens:", sum(HDB_train_sf$kindergartens_350m >= 5), "\n")
cat("Maximum kindergartens within 350m:", max(HDB_train_sf$kindergartens_350m), "\n")
cat("Average kindergartens within 350m:", round(mean(HDB_train_sf$kindergartens_350m), 2), "\n")

# Show distribution
cat("\nKindergarten count distribution:\n")
kg_counts <- table(HDB_train_sf$kindergartens_350m)
for(i in 1:min(10, length(kg_counts))) {
  cat("  ", names(kg_counts)[i], "kindergartens:", kg_counts[i], "properties\n")
}

# Show sample of results
cat("\n=== Sample Kindergartens Proximity ===\n")
kg_sample <- HDB_train_sf %>%
  select(block, street_name, town, kindergartens_350m) %>%
  st_drop_geometry() %>%
  arrange(desc(kindergartens_350m)) %>%
  head(10)
print(kg_sample)

# Save with kindergartens proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_kindergartens.rds")
cat("✓ Training data with kindergartens proximity saved\n")


```
### Number of Childcare Centres within 350m

```{r}
#  Childcare Centres within 350m
cat("===  Childcare Centres within 350m ===\n")

# Load the training data with previous proximity features
HDB_train_sf <- readRDS("data/rds/HDB_train_with_kindergartens.rds")

# Load childcare centres data
cat("Loading childcare centres data...\n")
if(file.exists("data/rawdata/ChildCareServices.geojson")) {
  childcare <- st_read("data/rawdata/ChildCareServices.geojson", quiet = TRUE) %>%
    st_transform(crs = 3414)
  cat("✓ Childcare centres data loaded from GeoJSON\n")
} else {
  stop("ChildCareServices.geojson not found!")
}

# Check childcare data
cat("Childcare centres found:", nrow(childcare), "\n")
cat("Childcare centres CRS:", st_crs(childcare)$input, "\n")

# Count childcare centres within 350m of each HDB
cat("Counting childcare centres within 350m...\n")
within_350m_cc <- st_is_within_distance(HDB_train_sf, childcare, dist = 350)
count_cc_350m <- lengths(within_350m_cc)

# Add count to dataset
HDB_train_sf <- HDB_train_sf %>%
  mutate(childcare_350m = count_cc_350m)

# Display results
cat("✓ Childcare centres proximity analysis completed\n")
cat("\n=== Childcare Centres within 350m Summary ===\n")
cat("Properties with 0 childcare centres:", sum(HDB_train_sf$childcare_350m == 0), "\n")
cat("Properties with 1+ childcare centres:", sum(HDB_train_sf$childcare_350m >= 1), "\n")
cat("Properties with 2+ childcare centres:", sum(HDB_train_sf$childcare_350m >= 2), "\n")
cat("Properties with 5+ childcare centres:", sum(HDB_train_sf$childcare_350m >= 5), "\n")
cat("Maximum childcare centres within 350m:", max(HDB_train_sf$childcare_350m), "\n")
cat("Average childcare centres within 350m:", round(mean(HDB_train_sf$childcare_350m), 2), "\n")

# Show distribution
cat("\nChildcare centres count distribution:\n")
cc_counts <- table(HDB_train_sf$childcare_350m)
for(i in 1:min(10, length(cc_counts))) {
  cat("  ", names(cc_counts)[i], "childcare centres:", cc_counts[i], "properties\n")
}

# Show sample of results
cat("\n=== Sample Childcare Centres Proximity ===\n")
cc_sample <- HDB_train_sf %>%
  select(block, street_name, town, childcare_350m) %>%
  st_drop_geometry() %>%
  arrange(desc(childcare_350m)) %>%
  head(10)
print(cc_sample)

# Save with childcare centres proximity
saveRDS(HDB_train_sf, "data/rds/HDB_train_with_childcare.rds")
cat("✓ Training data with childcare centres proximity saved\n")


```

### Numbers of primary school within 1km

```{r}
# Primary Schools within 1km
cat("===  Primary Schools within 1km ===\n")

# Load the training data with all previous location factors
HDB_train_sf <- readRDS("data/rds/HDB_train_all_location_factors.rds")

# Load preschools data (which includes primary schools)
cat("Loading preschools data for primary schools count...\n")
preschool <- st_read("data/rawdata/PreSchoolsLocation.kml", quiet = TRUE) %>% 
  st_transform(crs = 3414)

cat("✓ Preschools data loaded\n")
cat("Number of preschool/primary school locations:", nrow(preschool), "\n")
cat("Data CRS:", st_crs(preschool)$input, "\n")

# Count primary schools within 1km of each HDB
cat("Counting primary schools within 1km...\n")
within_1km_ps <- st_is_within_distance(HDB_train_sf, preschool, dist = 1000)
count_ps_1km <- lengths(within_1km_ps)

# Add count to dataset (using consistent naming)
HDB_train_sf <- HDB_train_sf %>%
  mutate(prischool_1km = count_ps_1km)

# Display results
cat("✓ Primary schools within 1km analysis completed\n")
cat("\n=== Primary Schools within 1km Summary ===\n")
cat("Properties with 0 primary schools:", sum(HDB_train_sf$prischool_1km == 0), "\n")
cat("Properties with 1+ primary schools:", sum(HDB_train_sf$prischool_1km >= 1), "\n")
cat("Properties with 5+ primary schools:", sum(HDB_train_sf$prischool_1km >= 5), "\n")
cat("Properties with 10+ primary schools:", sum(HDB_train_sf$prischool_1km >= 10), "\n")
cat("Maximum primary schools within 1km:", max(HDB_train_sf$prischool_1km), "\n")
cat("Average primary schools within 1km:", round(mean(HDB_train_sf$prischool_1km), 2), "\n")

# Show distribution
cat("\nPrimary schools count distribution:\n")
ps_counts <- table(HDB_train_sf$prischool_1km)
for(i in 1:min(15, length(ps_counts))) {
  cat("  ", names(ps_counts)[i], "primary schools:", ps_counts[i], "properties\n")
}

# Show sample of results
cat("\n=== Sample Primary Schools Proximity ===\n")
ps_sample <- HDB_train_sf %>%
  select(block, street_name, town, prischool_1km) %>%
  st_drop_geometry() %>%
  arrange(desc(prischool_1km)) %>%
  head(10)
print(ps_sample)

# Save the COMPLETE dataset with ALL location factors
saveRDS(HDB_train_sf, "data/rds/HDB_train_complete_location_factors.rds")
cat("✓ COMPLETE training data with ALL location factors saved\n")


```

## Correlation Analysis and Train-Test Split

```{r}
# Step 9: Correlation Analysis and Data Preparation
cat("=== Step 9: Correlation Analysis and Data Preparation ===\n")

# Load required packages
library(dplyr)
library(sf)
library(corrplot)
library(car)

# Load the complete training data with all location factors
HDB_train_complete <- readRDS("data/rds/HDB_train_complete_location_factors.rds")

# 1. Convert to non-spatial for correlation analysis
hdb_train_nogeo <- HDB_train_complete %>% st_drop_geometry()

# 2. Select only numeric predictors for correlation
cat("Selecting numeric predictors for correlation analysis...\n")
numeric_predictors <- hdb_train_nogeo %>%
  select(where(is.numeric)) %>%
  select(-LATITUDE, -LONGITUDE)  # Remove coordinates

cat("Numeric predictors selected:", ncol(numeric_predictors), "variables\n")
print(names(numeric_predictors))

# 3. Compute correlation matrix
cat("\nComputing correlation matrix...\n")
corr_matrix <- cor(numeric_predictors, use = "complete.obs")

# 4. Plot correlation matrix
cat("Plotting correlation matrix...\n")
corrplot(corr_matrix,
         diag = FALSE,
         order = "AOE",
         tl.pos = "td",
         tl.cex = 0.5,
         method = "color",
         type = "upper",
         title = "Correlation Matrix of Predictors",
         mar = c(0, 0, 2, 0))

# 5. Identify highly correlated variables (> 0.8)
cat("\nIdentifying highly correlated variables (r > 0.8)...\n")
high_corr <- which(abs(corr_matrix) > 0.8 & upper.tri(corr_matrix), arr.ind = TRUE)
if(length(high_corr) > 0) {
  high_corr_pairs <- data.frame(
    Var1 = rownames(corr_matrix)[high_corr[, 1]],
    Var2 = colnames(corr_matrix)[high_corr[, 2]],
    Correlation = corr_matrix[high_corr]
  ) %>%
    arrange(desc(abs(Correlation)))
  
  print(high_corr_pairs)
} else {
  cat("No highly correlated variable pairs found (r > 0.8)\n")
}

# 6. Check for multicollinearity using VIF
cat("\nChecking Variance Inflation Factor (VIF)...\n")

# Create a temporary model to check VIF
predictor_names <- setdiff(names(numeric_predictors), "resale_price")
temp_formula <- as.formula(paste("log(resale_price) ~", paste(predictor_names, collapse = " + ")))

temp_model <- lm(temp_formula, data = hdb_train_nogeo)
vif_results <- vif(temp_model)

cat("VIF Results:\n")
print(vif_results)

# 7. Identify variables to potentially remove (VIF > 10)
high_vif <- which(vif_results > 10)
if(length(high_vif) > 0) {
  cat("\nVariables with high VIF (>10):\n")
  print(vif_results[high_vif])
} else {
  cat("\nNo variables with high VIF (>10) detected\n")
}

# 8. Create final modeling dataset
cat("\nCreating final modeling dataset...\n")
# Remove geometry and select relevant columns
hdb_train_final <- HDB_train_complete %>%
  st_drop_geometry() %>%
  select(-month, -town, -flat_type, -block, -street_name, -flat_model, -LATITUDE, -LONGITUDE)

cat("Final training dataset dimensions:", dim(hdb_train_final), "\n")
cat("Predictors:", ncol(hdb_train_final) - 1, "variables\n")
cat("Response variable: resale_price\n")

# 9. Save the final modeling dataset
saveRDS(hdb_train_final, "data/rds/hdb_train_modeling.rds")
cat("✓ Final modeling dataset saved: data/rds/hdb_train_modeling.rds\n")

# Summary
cat("\n=== CORRELATION ANALYSIS SUMMARY ===\n")
cat("✓ Correlation matrix computed and plotted\n")
cat("✓ VIF analysis completed\n")
cat("✓ High correlation pairs identified\n")
cat("✓ Final modeling dataset prepared\n")
cat("✓ Ready for model building\n")

```

