---
title: "Take home Exercise 3 : Modelling HDB Resale Prices with Geographically Weighted Methods"
author: "Priya"
date: "November 4, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
format: 
  html:
    code-fold: false
    code-summary: "code chunk"
    number-sections: true
    code-overflow: scroll
    number-depth: 4
    code-line-numbers: false
---

## Overview

This study develops predictive models for HDB 4-room resale prices in Singapore using geographically weighted methods. The analysis compares traditional linear regression, geographically weighted regression (GWR), and random forest approaches to identify the most accurate predictive model for housing prices from July to September 2025 based on data from July 2024 to June 2025.

## Loading Packages

```{r}
pacman::p_load(httr, sf, jsonlite, progress, 
               tmap, rsample, Metrics, tidyverse, sf, spdep, GWmodel, SpatialML,
               knitr, kableExtra, spatialRF,
               randomForestExplainer, plotly,
               corrplot, ggspatial, units, caret)
```

## Data Preparation
Loading and Temporal Filtering

```{r}
#| label: load-data

# Training data: July 2024 - June 2025
train_raw <- read_csv("data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2024-07" & month <= "2025-06")

# Test data: July 2025 - September 2025  
test_raw <- read_csv("data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2025-07" & month <= "2025-09")

cat("Training data points:", nrow(train_raw), "\n")
cat("Test data points:", nrow(test_raw), "\n")

```

## Data Cleaning and Feature Engineering

```{r}
#| label: data-cleaning

clean_street_names <- function(data) {
  data %>%
    mutate(street_name = gsub("ST\\.", "SAINT", street_name))
}

engineer_features <- function(data) {
  data %>%
    mutate(
      # Convert storey range to ordinal scale
      storey_order = case_when(
        storey_range == "01 TO 03" ~ 1,
        storey_range == "04 TO 06" ~ 2,
        storey_range == "07 TO 09" ~ 3,
        storey_range == "10 TO 12" ~ 4,
        storey_range == "13 TO 15" ~ 5,
        storey_range == "16 TO 18" ~ 6,
        storey_range == "19 TO 21" ~ 7,
        storey_range == "22 TO 24" ~ 8,
        storey_range == "25 TO 27" ~ 9,
        storey_range == "28 TO 30" ~ 10,
        storey_range == "31 TO 33" ~ 11,
        storey_range == "34 TO 36" ~ 12,
        storey_range == "37 TO 39" ~ 13,
        storey_range == "40 TO 42" ~ 14,
        TRUE ~ 7  # Default for unknown ranges
      ),
      # Calculate remaining lease in months
      lease_commence_year = as.numeric(lease_commence_date),
      current_year = 2025,
      flat_age = current_year - lease_commence_year,
      remaining_lease_years = 99 - flat_age,
      remaining_lease_mths = remaining_lease_years * 12
    ) %>%
    select(-lease_commence_year, -current_year, -flat_age, -remaining_lease_years)
}

# Apply cleaning and feature engineering
train_clean <- train_raw %>% clean_street_names() %>% engineer_features()
test_clean <- test_raw %>% clean_street_names() %>% engineer_features()
```

Feature engineering transforms categorical variables (storey_range) into ordinal numeric variables and calculates remaining lease in months, which are more suitable for regression models.

## Geocoding with Error Handling

```{r}
#| label: improved-geocoding

# Improved geocoding function with error handling
geocode_safe <- function(block, streetname, max_retries = 3) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address,
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  for (attempt in 1:max_retries) {
    tryCatch({
      res <- GET(base_url, query = query)
      if (res$status_code == 200) {
        restext <- content(res, as = "text")
        output <- fromJSON(restext)
        
        if (!is.null(output$results) && nrow(output$results) > 0) {
          return(data.frame(
            LATITUDE = as.numeric(output$results$LATITUDE[1]),
            LONGITUDE = as.numeric(output$results$LONGITUDE[1])
          ))
        }
      }
    }, error = function(e) {
      if (attempt == max_retries) {
        warning(paste("Failed to geocode:", address))
        return(data.frame(LATITUDE = NA, LONGITUDE = NA))
      }
      Sys.sleep(0.5) # Wait before retry
    })
  }
  return(data.frame(LATITUDE = NA, LONGITUDE = NA))
}

# Sample data for computational efficiency
set.seed(1234)
train_sample <- train_clean %>% sample_n(1000)  # Increased sample size
test_sample <- test_clean %>% sample_n(500)    

# Geocode training data
train_sample$LATITUDE <- NA
train_sample$LONGITUDE <- NA

cat("Geocoding training data...\n")
pb <- progress_bar$new(total = nrow(train_sample))
for(i in 1:nrow(train_sample)) {
  temp_output <- geocode_safe(train_sample$block[i], train_sample$street_name[i])
  train_sample$LATITUDE[i] <- temp_output$LATITUDE
  train_sample$LONGITUDE[i] <- temp_output$LONGITUDE
  pb$tick()
}

# Geocode test data
test_sample$LATITUDE <- NA
test_sample$LONGITUDE <- NA

cat("Geocoding test data...\n")
pb <- progress_bar$new(total = nrow(test_sample))
for(i in 1:nrow(test_sample)) {
  temp_output <- geocode_safe(test_sample$block[i], test_sample$street_name[i])
  test_sample$LATITUDE[i] <- temp_output$LATITUDE
  test_sample$LONGITUDE[i] <- temp_output$LONGITUDE
  pb$tick()
}

# Remove failed geocoding attempts
train_sample <- train_sample %>% filter(!is.na(LATITUDE))
test_sample <- test_sample %>% filter(!is.na(LATITUDE))

cat("Successfully geocoded training samples:", nrow(train_sample), "\n")
cat("Successfully geocoded test samples:", nrow(test_sample), "\n")
```
;



